---
title: "Seurat v5 + 10x VDJ v2 Integration"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Seurat v5 + 10x VDJ v2 Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = TRUE,
  warning = TRUE
)
```

This vignette shows how to:
- Start from a Seurat v5 object.
- Attach 10x 5' VDJ v2 outputs to the object with flexible cell-name mapping.
- Inspect mapping diagnostics and basic metadata written by vdjmatchR.

Later sections (placeholders) will add pairing/QC, clonotype definitions, and vdjmatchR matching.

Note: paths below reference user-specific data; code chunks are guarded to run only if files exist.

## Prerequisites

```{r packages}
library(Seurat)       # v5
library(vdjmatchR)
library(magrittr)
```

## Provide your inputs

Set the path to a Seurat v5 object (RDS) and to the 10x VDJ directory (`vdj_t`).

```{r params, eval=TRUE}
# Edit these to your environment
seurat_rds <- "/Volumes/fh/fast/furlan_s/user/jperalta/MMCarT_240212_reseq/cds/mmcart_seu_viewmastR_CARpositive_251023.RDS"

# Example 10x VDJ v2 directory (replace as needed)
vdj_dir <- "/Volumes/fh/fast/furlan_s/user/jperalta/MMCarT_240212_reseq/FOLDERNAME1/data/FOLDERNAME2/outs/per_sample_outs/FOLDERNAME2/vdj_t"

```

## Load Seurat object

```{r load-seurat, eval=TRUE}
if (nzchar(seurat_rds) && file.exists(seurat_rds)) {
  orig <- readRDS(seurat_rds)
} else {
  message("Set SEURAT_RDS_PATH to a Seurat v5 RDS file to run this vignette end-to-end.")
}

prefixes <- sub("[ACGT]{16}-\\d+$", "", Cells(orig)) %>% unique()
folders2 <- sub("^[^_]+_(.+)_$", "\\1", prefixes)
folders1 <- sub("_[^_]+$", "", folders2)
matchdf <- data.frame(PREFIXES=prefixes, FOLDERNAME1=folders1, FOLDERNAME2=folders)

```

## Attach 10x VDJ v2 to Seurat

The function `vdj_attach_10x_vdj_v2()` reads `filtered_contig_annotations.csv` and `clonotypes.csv`,
maps 10x barcodes to Seurat cell names (with flexible strategies), and stores heavy tables in
`obj@tools$vdj`. Minimal metadata (10x clonotype id/size) can be written per cell.

```{r attach, eval=TRUE}

obj <- vdj_attach_10x_vdj_v2_batch(
  orig,
  matchdf,
  vdj_dir_template = vdj_dir)

obj$vdj.cdr3_aa_both <- paste0(obj$vdj.tra.cdr3_aa, "_", obj$vdj.trb.cdr3_aa)

data.frame(old=obj@meta.data[obj$vdj.cdr3_aa_both != obj$cdr3,]$cdr3, new=obj@meta.data[obj$vdj.cdr3_aa_both != obj$cdr3,]$vdj.cdr3_aa_both)


table(obj$vdj.cdr3_aa_both != obj$cdr3)

obj$nomatch <- obj$vdj.cdr3_aa_both != obj$cdr3
DimPlot(obj, group.by = "nomatch")
```


### Inspect attached tables and metadata

```{r inspect, eval=TRUE}
if (exists("obj") && !is.null(obj@tools$vdj$contigs)) {
  cat("Contigs rows:", nrow(obj@tools$vdj$contigs), "\n")
  print(utils::head(obj@tools$vdj$contigs[ , c("barcode","chain","v_gene","j_gene","cdr3","umis","cell","sample_id")], 5))
}

if (exists("obj")) {
  cols <- intersect(c("vdj.clone_id_10x","vdj.clone_size_10x"), colnames(obj@meta.data))
  if (length(cols)) print(utils::head(obj@meta.data[, cols, drop = FALSE]))
}
```

## Next steps (placeholders)

The following functions will be added in subsequent phases of development:

- `vdj_collapse_pairs_seurat(obj, keep_two_alpha = TRUE, drop_multi_trb = TRUE, ...)`:
  - Applies QC (productive, high_confidence, full_length; min UMIs/reads).
  - Selects one TRB (drop cells with >1 TRB) and up to two TRA.
  - Writes per-cell pairing metadata (`vdj.*`) and stores a `pairs` table in `tools$vdj`.

- `vdj_define_clonotypes_seurat(obj, mode = c("tenx","cdr3_aa","cdr3aa_vj"))`:
  - Default: use 10x clonotype IDs.
  - Optionally recompute clones by exact AA sequences (± V/J).

- `vdj_match_seurat(db, obj, chain = "TRB", scope = "0,0,0,0", top_n = 1L, ...)`:
  - Builds TRB queries from Seurat metadata/pairs and calls `match_tcr_many_df`.
  - Stores full results in `tools$vdj$hits_trb` and per-cell summaries in metadata (`vdjdb.*`).

## Match Seurat TCRs to VDJdb

This section shows how to take the TRA/TRB sequences written into Seurat metadata
and match them against VDJdb, producing a compact table of each cell's top hit per chain.

```{r vdjdb-match, eval=TRUE}
if (exists("obj")) {
  # 1) Load a VDJdb and filter to human and each chain of interest
  db_path <- vdjdb_packaged_path(use_fat_db = FALSE)
  if (nzchar(db_path) && file.exists(db_path)) {
    db <- RDatabase$new_from_file(db_path)

    # Helper to build queries from metadata and get top-1 hit per cell
    get_top_hits <- function(obj, chain = c("TRB", "TRA"), scope = "0,0,0,0") {
      chain <- match.arg(chain)
      ch <- tolower(chain)
      md <- obj@meta.data
      cdr3_col <- paste0("vdj.", ch, ".cdr3_aa")
      v_col    <- paste0("vdj.", ch, ".v_gene")
      j_col    <- paste0("vdj.", ch, ".j_gene")
      if (!all(c(cdr3_col, v_col, j_col) %in% colnames(md))) return(NULL)
      keep <- which(!is.na(md[[cdr3_col]]) & nzchar(md[[cdr3_col]]))
      if (!length(keep)) return(NULL)
      queries <- data.frame(
        cell = rownames(md)[keep],
        cdr3 = as.character(md[[cdr3_col]][keep]),
        v    = as.character(md[[v_col]][keep]),
        j    = as.character(md[[j_col]][keep]),
        stringsAsFactors = FALSE
      )
      # Filter DB to species/gene
      db_chain <- filter_db(db, species = "HomoSapiens", gene = chain, min_vdjdb_score = 0)
      hits <- match_tcr_many_df(db_chain, queries$cdr3, queries$v, queries$j, scope = scope, top_n = 1L)
      if (!nrow(hits)) return(NULL)
      # Attach cell back by query_index and return a compact table
      hits$cell <- queries$cell[as.integer(hits$query_index)]
      # Keep key columns commonly used from VDJdb; adjust as needed
      keep_cols <- intersect(c("cell","query_index","cdr3","v_segment","j_segment","gene","epitope","antigen.species","vdjdb_score","source"), colnames(hits))
      hits[, keep_cols, drop = FALSE]
    }

    top_trb <- get_top_hits(obj, chain = "TRB", scope = "0,0,0,0")
    top_tra <- get_top_hits(obj, chain = "TRA", scope = "0,0,0,0")

    if (!is.null(top_trb)) {
      cat("Top TRB hits (first 10):\n")
      print(utils::head(top_trb, 10))
    }
    if (!is.null(top_tra)) {
      cat("\nTop TRA hits (first 10):\n")
      print(utils::head(top_tra, 10))
    }
  } else {
    message("VDJdb packaged path not found; set a user DB via vdjdb_set_user_db().")
  }
}
```

## Session info

```{r session}
sessionInfo()
```
